---
title: "Time Series Analysis - Practical 5"
subtitle: "Commercial Bank Real Estate Loans Analysis"
author: "TSA Practical 5"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)
```

# Objective

Analyze the monthly volume of commercial bank real estate loans (in billions of dollars) to:

- Import and visualize the data
- Identify dominant components (trend, seasonality, etc.)
- Test for stationarity using ACF/PACF plots
- Perform Augmented Dickey-Fuller (ADF) test for stationarity

# Dataset Information

- **File**: `bank_case.txt`
- **Description**: Monthly volume of commercial bank real estate loans in billions of dollars
- **Number of observations**: 70 months

---

# (a) Import Data into R Environment

Read the data from the text file and display basic information:

```{r import-data}
# Read the data from text file
bank_data <- scan("bank_case.txt", quiet = TRUE)

# Display the first few values
cat("First 10 values:\n")
print(head(bank_data, 10))

cat("\nLast 10 values:\n")
print(tail(bank_data, 10))

cat("\nTotal number of observations:", length(bank_data), "\n")
cat("Range: [", min(bank_data), ",", max(bank_data), "]\n")
cat("Mean:", mean(bank_data), "\n")
cat("Standard Deviation:", sd(bank_data), "\n")
```

---

# (b) Convert Data into Time Series Object

Create a time series object with monthly frequency:

```{r create-ts}
# Convert to time series object
# Assuming monthly data (frequency = 12)
bank_ts <- ts(bank_data, frequency = 12, start = c(1, 1))

# Display time series summary
print(bank_ts)
```

---

# (c) Plot Data to Identify Dominant Component

## Main Time Series Plot

```{r plot-main, fig.cap="Commercial Bank Real Estate Loans Over Time"}
# Main time series plot with trend line
plot(bank_ts, 
     main = "Commercial Bank Real Estate Loans Over Time",
     xlab = "Time (Months)",
     ylab = "Loan Volume (Billions of Dollars)",
     col = "blue",
     lwd = 2)
grid()

# Add a trend line
abline(lm(bank_ts ~ time(bank_ts)), col = "red", lwd = 2, lty = 2)
legend("topleft", 
       legend = c("Original Data", "Trend Line"), 
       col = c("blue", "red"), 
       lty = c(1, 2), 
       lwd = 2)
```

**Observation**: The plot shows a clear **upward trend** in the loan volume over time, with values increasing from approximately 46.5 to 87.6 billion dollars.

## Time Series Decomposition

```{r decompose, fig.height=8, fig.cap="Decomposition of Time Series Components"}
# Decompose the time series
decomposed <- decompose(bank_ts)
plot(decomposed)
```

**Analysis of Components**:

- **Trend**: Strong upward trend throughout the series
- **Seasonal**: Some periodic fluctuations
- **Random**: Residual variation after removing trend and seasonal components

**Conclusion**: The **dominant component is the TREND**, showing consistent growth in bank real estate loans.

---

# (d) Check Stationarity Using ACF/PACF Plots

```{r acf-pacf, fig.height=8, fig.cap="ACF and PACF Plots for Stationarity Assessment"}
# Set up plotting area
par(mfrow = c(2, 1))

# ACF plot
acf(bank_ts, 
    main = "Autocorrelation Function (ACF)",
    lag.max = 36,
    col = "blue",
    lwd = 2)

# PACF plot
pacf(bank_ts, 
     main = "Partial Autocorrelation Function (PACF)",
     lag.max = 36,
     col = "red",
     lwd = 2)

# Reset plotting area
par(mfrow = c(1, 1))
```

## ACF/PACF Interpretation

- **ACF**: Shows slow, gradual decay - this is a strong indicator of **non-stationarity**
- **PACF**: First few lags are significant, then cuts off
- **Conclusion**: The ACF pattern (slow decay) strongly suggests the series is **non-stationary**

**Why this indicates non-stationarity**:

- Stationary series would show ACF cutting off quickly
- The persistent high correlation at many lags indicates a trend component
- Non-constant mean over time (upward trend visible in plot)

---

# (e) Augmented Dickey-Fuller (ADF) Test

```{r adf-test}
# Load tseries package for ADF test
library(tseries, quietly = TRUE)

cat("\n========== Augmented Dickey-Fuller Test ==========\n")
cat("Null Hypothesis (H0): Series has a unit root (non-stationary)\n")
cat("Alternative Hypothesis (H1): Series is stationary\n")
cat("Significance level: α = 0.05\n\n")

# Perform ADF test
adf_test <- adf.test(bank_ts, alternative = "stationary")
print(adf_test)
```

## Test Interpretation

```{r adf-interpretation, echo=FALSE}
cat("\n--- Interpretation ---\n")
cat("Test Statistic:", adf_test$statistic, "\n")
cat("P-value:", adf_test$p.value, "\n\n")

if (adf_test$p.value < 0.05) {
    cat("Decision: REJECT H0 (p-value < 0.05)\n")
    cat("Conclusion: The series is STATIONARY\n")
} else {
    cat("Decision: FAIL TO REJECT H0 (p-value >= 0.05)\n")
    cat("Conclusion: The series is NON-STATIONARY\n")
    cat("\nImplication: The series contains a unit root and exhibits non-stationary behavior.\n")
    cat("Recommendation: Apply differencing to achieve stationarity.\n")
}
```

---

# First Difference Analysis

Since the original series is non-stationary, we apply first differencing:

```{r first-difference}
cat("\n========== Testing First Difference ==========\n\n")

# Compute first difference
bank_diff <- diff(bank_ts)

cat("Summary statistics of differenced series:\n")
cat("Mean:", mean(bank_diff), "\n")
cat("Variance:", var(bank_diff), "\n")
cat("Standard Deviation:", sd(bank_diff), "\n")
```

## Plot of First Difference

```{r plot-diff, fig.cap="First Difference of Bank Loan Data"}
plot(bank_diff,
     main = "First Difference of Bank Loan Data",
     xlab = "Time",
     ylab = "Differenced Values (Billions $)",
     col = "darkgreen",
     lwd = 2)
grid()
abline(h = 0, col = "red", lty = 2)
abline(h = mean(bank_diff), col = "blue", lty = 2)
legend("topright", 
       legend = c("Differenced Series", "Zero Line", "Mean"), 
       col = c("darkgreen", "red", "blue"), 
       lty = c(1, 2, 2), 
       lwd = c(2, 1, 1))
```

**Observation**: The differenced series appears more stable, fluctuating around a constant mean.

## ADF Test on First Difference

```{r adf-diff}
# ADF test on differenced series
adf_diff <- adf.test(bank_diff, alternative = "stationary")
print(adf_diff)

cat("\n--- Interpretation of Differenced Series ---\n")
if (adf_diff$p.value < 0.05) {
    cat("Decision: REJECT H0 (p-value < 0.05)\n")
    cat("Conclusion: The FIRST DIFFERENCE is STATIONARY\n")
    cat("\nThe original series is integrated of order 1, denoted I(1)\n")
    cat("This means one differencing operation is sufficient to achieve stationarity.\n")
} else {
    cat("Decision: FAIL TO REJECT H0\n")
    cat("Conclusion: Additional differencing may be needed\n")
}
```

---

# Summary of Findings

## Key Results

1. **Data Characteristics**:
   - 70 monthly observations
   - Range: 46.5 to 87.6 billion dollars
   - Strong upward trend throughout the period

2. **Dominant Component**: 
   - **TREND** - Clear, consistent upward movement in loan volumes

3. **Stationarity Tests**:
   
   | Test Method | Result | Evidence |
   |-------------|--------|----------|
   | Visual Inspection | Non-stationary | Upward trend visible |
   | ACF Plot | Non-stationary | Slow decay of autocorrelations |
   | ADF Test | Non-stationary | p-value = `r round(adf_test$p.value, 4)` > 0.05 |
   | First Difference ADF | Check results | p-value = `r round(adf_diff$p.value, 4)` |

4. **Statistical Conclusion**:
   - Original series is **non-stationary** (fails ADF test)
   - Contains a unit root
   - Mean and variance change over time
   - First differencing applied to achieve stationarity

## Implications for Time Series Modeling

- **ARIMA Modeling**: The series requires differencing (d=1 or d=2)
- **Forecast Method**: Should account for the trend component
- **Transformation**: First difference removes the trend, making the series more suitable for modeling
- **Model Suggestion**: Consider ARIMA(p,1,q) model

---

# Theoretical Background

## Stationarity

A time series $\{X_t\}$ is **stationary** if:

1. $E[X_t] = \mu$ (constant mean)
2. $Var(X_t) = \sigma^2$ (constant variance)
3. $Cov(X_t, X_{t+k}) = \gamma_k$ (covariance depends only on lag $k$)

## Augmented Dickey-Fuller Test

The ADF test examines the null hypothesis:

$$H_0: \text{The series has a unit root (non-stationary)}$$
$$H_1: \text{The series is stationary}$$

The test equation is:
$$\Delta Y_t = \alpha + \beta t + \gamma Y_{t-1} + \sum_{i=1}^{p} \delta_i \Delta Y_{t-i} + \epsilon_t$$

Where $\gamma < 0$ under the alternative hypothesis.

## Differencing Operator

First difference: $\nabla X_t = X_t - X_{t-1} = (1-B)X_t$

Where $B$ is the backshift operator: $BX_t = X_{t-1}$

---

**Analysis Complete**
